A conditional permutation-based approach to test confounder effect and center-bias in machine learning models

### 2021-08-10 manuscript.tex
polished

\begin{abstract} % 150 words
The lack of rigorous non-parametric statistical tests of confoudner-effects significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recently established theoretical framework for permutation-based conditional independence testing and probe the null hypotheses of \emph{unbiased} and \emph{fully biased model}, respectively.
The proposed tests set no assumptions about the distribution of the predictive model output that is often non-normal.
Simulations show that the tests are applicable in a wide variety of cases with a remarkable statistical power.
Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset identified significantly biased predictions. Some of these biases were previously unreported and found to be hard to mitigate with state-of-the-art confound mitigation approaches.
The package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}} can aid the assessment of the generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
\end{abstract}

@@

In a predictive modelling setting, let $\y$ denote the target variable, $\yhat$ denote model output, i.e. the predictions for $\y$ and let $\c$ denote a variable which is considered as a confounder. Depending on the research question, the influence of $\c$ on the model predictions $\yhat$ is subject to be kept at a negligible level or it is to be proven that, at least, the model is not completely driven by $\c$. Below, I summarize how testing conditional independence \citep{dawid1979conditional} across these variables allows addressing the above questions and build on the conditional permutation test of \cite{berrett2020conditional} to perform inference without assumptions on the distribution of the model output, where normality is often violated \citep{garcia2009study, kristensen2017whole}.

\subsubsection*{Conditional independence}

Let us assume that $\y$, $\yhat$ and $\c$ consist of independent and identically distributed (IID) data points $(y_i, \hat{y}_i, c_i)$ for $i=1, \dots , n$ and write $\y=(y_1, \dots ,y_n)$, $\yhat=(\hat{y}_1, \dots, \hat{y}_n)$ and $\c=(c_1, \dots, c_n)$.

Obviously, a strong association between $\yhat$ and $\c$ can imply that the model is biased; it's predictions are driven by the confounder rather than information about the target variable.

Testing unconditional independence between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confoudner-bias in predictive modelling.
For instance, even if the null hypothesis $H0 : \yhat \independent \c$ is false, $\yhat$ and $\c$ might be only marginally dependent on $\c$, due to their dependence on $\y$. In other words, if the target variable $\y$ is associated to the confoudner variable $\c$, a model that is completely blind to $\c$ (and, therefore, unbiased) might still provide outputs $\yhat$ that are significantly associated with $\c$.

Our primary interest is, therefore, the \emph{conditionally independence} between $\yhat$ and $\c$ given $\y$ ($\yhat \independent \c | \y$) which, according to \cite{dawid1979conditional}, implies that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(X|Z)\mathbb{P}(Y|Z)$. The test with this null hypothesis will be referred to as the \emph{partial confounder test}.

Alternatively, one might be interested in testing $\yhat \independent \c | \y$. We will refer to the corresponding test as the \emph{full confoudner test}.

 Conditional independence - in its general form - lies at the heart of several fundamental concepts in statistics and  plays an increasingly important role in various fields of applied statistics, particularly in biomedical applications \citep{spirtes2000causation, peters2016causal, fiedler2011mediation, candes2016panning}. Recently, \cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
 Their "no free lunch" theorem implies that, without placing some assumptions on the joint distribution of $(\y, \yhat, \c)$, conditional Independence testing is effectively impossible. In other words, neither the full nor the partial confoudner tests can be constructed so that - for all distributions - they provides a valid type I error control and, at the same time, a non-trivial statistical power.

This result is in strong contrast to unconditional independence testing - where permutation tests  \citep{pitman1937significance, fisher1942189}, provide a general, distribution-free solution - and it has important implications for confounder testing in predictive modelling where the distribution of the model outputs - depending on the applied machine learning model - is unknown and often heavily non-normal.

Nevertheless, as shown by \cite{berrett2020conditional}, conditional independence testing is possible with information only about the conditional distribution of two out of the three variables.
In this case, sampling from a non-uniform distribution over the set of possible permutations $\pi$ of the one of the variables can incorporate available distributional information into permutation-based inference and achieve a valid type I error control and a non-trivial power.
Below we adapt the formalization of the procedure of \cite{berrett2020conditional} for the confoudner problem.

\subsubsection*{The conditional permutation test}

@@

The conditional permutation test (CPT) of \cite{berrett2020conditional} builds on the conditional randomization test of \cite{candes2016panning} and allows that \emph{no assumptions} are made about the distribution of one of the variables but, instead, it is assumed that the conditional distribution of the other two variables is known.

As many related papers, the work of \cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is the predictor(s) (features), $Y$ is the target variable and $Z$ is a potential confounder. In this setting, testing the null hypothesis $X \independent Y | Z$ aims to determine, whether $X$ affects $Y$, when controlling for $Z$.
For instance, in genome-wide association studies, CPT can be used o determine whether a particular genetic variant $X$ affects a response $Y$ such as disease status or some other phenotype, even after controlling for the rest of the genome, encoded in $Z$.

Here, different setting is considered, where supervised learning model is already fitted and we are focusing on model diagnostics, by testing the triplet $(\y,\yhat, \c)$, as introduced above. In such a setting, we are interested in how much variance the fitted model captures from the confounder $C$ or, in other words, to what extent the model is biased by the confounder variable $\c$ when predicting $\y$ with $\yhat$. The conceptual difference between the original and the proposed application of CPT is depicted on Fig. \ref{fig:schematic}.

@@

While option 3. is typically of interest when testing confounder bias of predictive models, option 1 might be also useful in diagnostics of predictive models, especially in the exploratory phase of model construction. Option 2 seems less appealing for model diagnostics and as detailed below, the proposed variety of the CPT framework does not allow forming a non-parametric method in this case.

Below, CPT is re-formalized for option 3, i.e. partial confoudner testing. The formulation of the full confoudner test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.

\subsubsection*{The partial confoudner test}

The partial confoudner test generates a null-distribution for an arbitrary predefined test statistic $T(\y,\yhat,\c)$ by sampling permutation based "copies" of $\c$,

@@

where, $Q(.|y)$ denotes the conditional distribution of $\c$ given $\y=y$ and $j=1,\dots, m$ indexes the "copy" of $\c$ so that $\c^{(i)} = (c_1^{(j)}, \dots, c_n^{(j)})$ is a permutation of the original vector $\c = (c_1, \dots, c_n)$. This mechanism creates copies $\c^{(1)}, \dots ,\c^{(m)}$ that are exchangeable with the original vector $\c$ under the null hypothesis that $\yhat \independent \c | \y$.

Under the null hypothesis, the triples $$(\y,\yhat,\c), (\y, \yhat, \c^{(1)}),\dots, (\y, \yhat, \c^{(m)})$$ are all identically distributed and exchangeable, and so are the 
$$T(\y,\yhat,\c), T(\y, \yhat, \c^{(1)}),\dots,T(\y, \yhat, \c^{(m)})$$
test statistics, as well.

Let $S_n$ denote the set of all permutations on the indices $\{1,\dots,n\}$ and $\c_\pi = (c_{\pi_1}, \dots, c_{\pi_n})$ the vector $\c$ with its elements reordered according to the permutation $\pi \in S_i$.
The permutation-based copies of $\c$ are then of $\c^{(j)} = \c_{\pi^{(j)}}$ which are drawn so that:

@@

While the required permutations could be theoretically sampled with a simple Metropolis-Hastings algorithm that draws uniformly from $S_n$ at random, this way the acceptance ratio would be extremely low, even for moderate $n$ (except there is very low dependence of $\c$ on $\y$), resulting in slow mixing times. \cite{berrett2020conditional} (Algorithm 1) propose a parallelized pairwise Markov-Chain Monte Carlo sampler for CPT, that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data. In their Theorem 2, \cite{berrett2020conditional} verify that the resulting Markov Chain yields the desired stationary distribution, even if the number of steps is small.

@@

Obtaining a relatively accurate, independent estimate of $D(\cdot|\y)$ (of any shape) for CPT inference is important. Berrett and colleagues recommend to use a large independent sample to obtain the log-likelihood matrix of the conditional distribution $D(\cdot|Z)$, which might be often viable in semi-supervised learning settings where unlabelled data $(X, Z)$ are easier to obtain than labelled data $(X, Y, Z)$.
Since in the proposed predictive modelling , normality might be often a reasonable assumption for $\y$ and $\c$ (but not for $\yhat$), it is possible to fall back to Gaussian assumptions and thereby eliminate the need for a large, independent sample to estimate $D(\cdot|\y)$.

@@

\begin{equation}
    \label{eq:linreg}
    \c = \alpha + \beta \y + \epsilon
\end{equation}

with the assumption that the residual error follows a Gaussian distribution and let $\boldsymbol{\mu} = \alpha + \beta \y$ and $\sigma^2 = E(\epsilon^2)$. Then, for the conditional Normal distribution $ (\c|\y=y_i) \sim \mathcal{N}\{\mu_i, \sigma^2\}$ and the log-likelihood, that is to be used in Eq. \ref{eq-pperm}, can be computed as follows:

@@

A straightforward choice of test statistic with the proposed Gaussian likelihood estimation is coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confoudner or classification):

@@

$$T(\y, \yhat, \c_\pi) = R^2(\yhat, \c_\pi)$$

@@

Using CPT to test confounder-bias in predictive modelling allows relaxing assumptions on $\hat{Y}$ (for both the partial and the full test) but - in line with the "no free lunch" theorem, requires knowing - or putting assumptions - on the joint distribution of the other two variables ($Y$ and $C$). 
\cite{berrett2020conditional} give a detailed analysis of the robustness of their CPT approach when estimating the conditional distribution with re-using the tested data and, also, against misspecifying the conditional distribution of interest.

Here I extend their results by performing validation that aids estimating the power of the test in a form that is accessible for power calculations in predictive modelling (considering various weights of the confounder and the target signals in $\hat{y}$ and in the case of varying levels of covariance between $Y$ and $C$).
Moreover, I perform a thorough investigation of the robustness of the test against the violation of normality (for both $(Y, C)$ and $\hat{Y}$).

@@

\subsubsection*{Gaussian Simulations}
The target variable $Y$ and the confounder variable $C$ were drawn randomly from a multivariate normal distribution:

$$ Y, C \sim \mathcal{N}(\mu_0, \Sigma_\sigma) $$,

@@

Next, we construct the simulated predicted values:

$$ \hat{Y} = w_Y Y + w_C C + w_e e$$

so that $w_Y + w_C + w_e = 1$, where $e \sim \mathcal{N}(0,1)$ is the noise component.

To test the implementation for categorical variables, simulated $Y$, $\hat{Y}$ and $C$ variables are binarized by thresholding at 0.

@@

$$ w_{X/Y} = w_C / w_Y $$

@@

$$ w_{C+Y} = w_C + w_Y $$,

@@

$$X' = sinh(\delta sinh^{-1}(X) - \epsilon)$$

@@

In the simulation procedure for the 'partial' confound test, 100 repetitions were performed of all combination of the following parameter values: 
\begin{itemize}
    \item covariance of $C$ and $Y$: $\sigma \in \{0, 0.2, 0.4, 0.6, 0.8\}$
    \item confound-to-target: $w_{C/Y} \in \{0, 0.1, 0.3, 0.5, 1\}$
    \item signal-in-prediction: $w_{C+Y} \in \{0, 0.3, 0.6, 0.9\}$
    \item sample size: $N \in \{50, 100, 500, 1000\}$
\end{itemize}

For testing 'full' confounder bias, identical parameters were used, except that instead of the 'confound-to-target' parameter $ w_{C/Y}$ we used the 'target-to-confound' parameter $ w_{Y/C} = w_Y / w_C $, but with the same parameter values $w_{Y/C} \in \{0, 0.1, 0.3, 0.5, 1\}$.

Next to the normally distributed case, 4 \emph{sinh-arcsinh} transformed cases were also investigated, with $\delta \in \{1, 1.05, 1.5, 5\}$ $\epsilon \in \{1, 3, 5, 10\}$

The 'partial' and 'full' confound tests, as implemented in version 0.7.1 of the package \emph{'mlconfound'} were run with default parameters, that is 1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations and by implying categorical variables, where needed.

@@

'Partial' and 'full' confound testing was used with default parameters (1000 permutations, 50 Markov-chain Monte Carlo steps) as implemented in version 0.7 of the package \emph{'mlconfound'}.

@@

Human Connectome Project contains imaging and behavioral data of more than 1000 healthy subjects \citep{van2013wu}. Preprocessed resting state fMRI connectivity data (partial correlation matrices) \citep{glasser2013minimal} as published with the HCP1200 release (N=999 participants with functional connectivity data) were used to build models that predict individual fluid intelligence scores (IQ), as measured with Penn Progressive Matrices \citep{duncan2000neural}.

@@

Features (functional concavities across 100 group independent component analysis based regions) were either (i) considered in their raw form or were subject to confound mitigation approaches by (ii) feature regression \citep{rao2017predictive} or (iii) COMBAT \citep{johnson2007adjusting, fortin2018harmonization}.
The feature mitigation strategies were separately applied for acquisition batch and age group as confounder variable.

The total of 5 types of features (raw, regressing out acquisition batch, regressing out age group, COMBAT with acquisition batch, COMBAT with age group) were independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the individual fluid intelligence scores with a ridge-regression \citep{hoerl1970ridge}. The $\alpha$ parameter of the ridge was considered as a hyperparameter ($\alpha \in \{0.000001, 0.01, 0.1, 1, 10, 100, 1000000\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with mean squared error as optimization metric. Confound mitigation was performed inside of the outer cross-validation loop, to avoid leakage.

Analysis code is available as jupyter notebook as part\footnote{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_hcp.ipynb} of the github repository of the 'mlconfound' package.

@@

The proposed method is applied to provide evidence of motion-bias in diagnostic predictive models of autism spectrum disorder (ASD), trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} involving 866 participants; ASD: 402, healthy control (HC): 464. Preprocessed regional timeseries data was obtained as shared with the paper \citep{dadi2019benchmarking} which was based on preprocessed image data provided by the Preprocessed Connectome Project \citep{craddock2013neuro}.

@@

The total of 5 types (raw, feature regression of site, feature regression of motion, COMBAT with site, COMBAT with motion) of features were independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the diagnosis (DX: ASD vs. HC) with a L2-regularized logistic regression (as e.g. in \citep{dadi2019benchmarking}). The $C$ parameter of the model was considered as a hyperparameter ($C \in \{0.1, 1, 10\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with mean squared error as optimization metric. Confound mitigation was performed inside of the outer cross-validation loop, to avoid leakage.

@@

The conditional permutation based 'partial' and 'full' confound tests for predictive modelling have been implemented in the python package \emph{'mlconfound'}. Simulations showed valid type I control and a high, practically relevant statistical power in a wide variety of data characteristics. Analysis of real functional brain connectivity data demonstrates the usefulness of the tests in guiding confound-mitigation strategies.

@@

Simulations based on normally distributed data justify that the proposed confounder tests provide valid type I error control (For detailed results, see first row and first column on Figure \ref{fig:sim-normal} for the 'partial' test on numerical data and Supplementary Figures \ref{fig:sim-ccc-full}, \ref{fig:sim-ccb-partial}, \ref{fig:sim-ccb-full}, \ref{fig:sim-bbc-partial}, \ref{fig:sim-bbc-full}, \ref{fig:sim-bbb-partial} and \ref{fig:sim-bbc-full} for the partial test categorical data and the full tests on both numeric and categorical data). The higher confidence limit for the type I error rates for one simulation case with 1
00 repetitions is 0.11 (approximately 5 simulation cases are expected to pass this limit across all 200 H0 simulations on one figure).
The results suggest that the proposed tests are somewhat overly conservative if the model has a high predictive performance.

Investigations of the tolerance to violation of normality assumptions showed that the proposed methods are remarkably robust to the violation of normality assumptions for the target variable $Y$ and the confounder $C$ (and completely non-parametric regarding the distribution of the predictions $\hat{y}$). The six simulation cases of normality violation , shown on Figure \ref{fig:sim-non-normal} involved various \emph{sinh-arcsinh} distributions, form totally normal (Fisher's kurtosis: 0, skewness) to severely skewed and heavy tailed (kurtosis: 42, skewness: 6).
The results show that the violation of normality - as expected - increases the type I error rate, but the increase - depending on the application - might not be severely problematic up to a kurtosis and skewness values as high as 5 and 2 respectively. Furthermore, normality violation becomes less problematic in case of high predictive performance (high association between the predicted and observed values).

@@

When trained on the raw (unadjusted) connectivity features, in all four cases, the confounders were statistically significantly associated to both the target variable and the model predictions (with weak-to-moderate effect sizes) and the proposed partial confounder test provided significant evidence against the null-hypothesis of no confounder bias (i.e. conditional independence between the confoudner and the predictions, given the target).

>>>>

\begin{abstract} % 150 words
The lack of rigorous non-parametric statistical tests of confounder-effects significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recently established theoretical framework for permutation-based conditional independence testing and probe the null hypotheses of \emph{unbiased} and \emph{fully biased model}, respectively.
The proposed tests set no assumptions about the distribution of the predictive model output that is often non-normal.
Simulations show that the tests are applicable in a wide variety of cases with a remarkable statistical power.
Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset identified significantly biased predictions. Some of these biases were previously unreported and found to be hard to mitigate with state-of-the-art confound mitigation approaches.
The package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}} can aid the assessment of the generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
\end{abstract}

@@

In a predictive modelling setting, let $\y$ denote the target variable, $\yhat$ denote model output, i.e. the predictions for $\y$ and let $\c$ denote a variable which is considered as a confounder (see Fig. \ref{fig:schematic}B). Let us assume, furthermore, that $\y$, $\yhat$ and $\c$ consist of independent and identically distributed (IID) data points $(y_i, \hat{y}_i, c_i)$ for $i=1, \dots , n$ so that $\y=(y_1, \dots ,y_n)$, $\yhat=(\hat{y}_1, \dots, \hat{y}_n)$ and $\c=(c_1, \dots, c_n)$.

Obviously, a strong association between $\yhat$ and $\c$ might indicate that the model is biased; it's predictions are driven by the confounder rather than information about the target variable.
Testing unconditional independence between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confounder-bias in predictive modelling.
For instance, even if the null hypothesis $H0 : \yhat \independent \c$ is false, $\yhat$ might be only marginally dependent on $\c$, due to their dependence on $\y$. In other words, if the target variable $\y$ is associated to the confounder variable $\c$, a model that is completely blind to $\c$ (and, therefore, unbiased) might still provide outputs $\yhat$ that are significantly associated with $\c$.

 Below, I summarize how testing conditional independence \citep{dawid1979conditional} with conditional permutation test \citep{berrett2020conditional} allows evaluating confounder-bias in predictive modelling, without placing assumptions on the distribution of the model output, where normality is often violated \citep{garcia2009study, kristensen2017whole}.

\subsubsection*{Conditional independence for testing confounder bias}

Depending on the research question, the influence of $\c$ on the model predictions $\yhat$ is subject to be kept at a negligible level or it is to be proven that, at least, the model is not completely driven by $\c$.

Our primary interest is, therefore, the \emph{conditionally independence} between $\yhat$ and $\c$ given $\y$ (written as $\yhat \independent \c | \y$) which, according to \cite{dawid1979conditional}, implies that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(\yhat|\y)\mathbb{P}(\c|\y)$. The test with this null hypothesis will be referred to as the \emph{partial confounder test}.

Alternatively, one might be interested in testing $\yhat \independent \c | \y$. We will refer to the corresponding test as the \emph{full confounder test}.

 Conditional independence - in its general form - lies at the heart of several fundamental concepts in statistics and  plays an increasingly important role in various fields of applied statistics, particularly in biomedical applications \citep{spirtes2000causation, peters2016causal, fiedler2011mediation, candes2016panning}. Recently, \cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
 Their "no free lunch" theorem implies that, without placing some assumptions on the joint distribution of $(\y, \yhat, \c)$, conditional Independence testing is effectively impossible. In other words, neither the full nor the partial confounder tests can be constructed so that - for all distributions - they provides a valid type I error control and, at the same time, a non-trivial statistical power.

This result is in strong contrast to unconditional independence testing - where permutation tests  \citep{pitman1937significance, fisher1942189}, provide a general, distribution-free solution - and it has important implications for confounder testing in predictive modelling where the distribution of the model outputs, depending on the applied machine learning model, is unknown and often heavily non-normal.

Nevertheless, as shown by \cite{berrett2020conditional}, conditional independence testing is possible with information only about the conditional distribution of two out of the three variables.
The conditional permutation test (CPT) of Berret and colleagues builds on the conditional randomization test of \cite{candes2016panning} and samples from a non-uniform distribution over the set of possible permutations $\pi$ of one of the variables, so that it incorporates available distributional information into permutation-based inference.

@@

As many related papers, the work of \cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is the predictor(s) (features), $Y$ is the target variable and $Z$ is a potential confounder. In this setting, testing the null hypothesis $X \independent Y | Z$ aims to determine, whether $X$ affects $Y$, when controlling for $Z$.
For instance, in genome-wide association studies, CPT can be used to determine whether a particular genetic variant $X$ affects a response $Y$ such as disease status or some other phenotype, even after controlling for the rest of the genome, encoded in $Z$.

In this paper, a different setting is considered, where the supervised learning model is already fitted and we are focusing on model diagnostics, by testing the triplet $(\y,\yhat, \c)$. The conceptual difference between the original and the proposed application of CPT is depicted on Fig. \ref{fig:schematic}.

@@

Option 3, i.e. partial confounder testing is typically of interest when testing confounder bias of predictive models. Option 1, i.e. full confounder testing, might be also useful in diagnostics of predictive models, especially in the exploratory phase of model construction. Option 2 seems less appealing for model diagnostics and importantly, the CPT framework does not allow constructing a test which is non-parametric on $\yhat$.

Below, CPT is adapted for partial confounder testing (option 3). The formulation of the full confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.

\subsubsection*{The partial confounder test}

The partial confounder test generates a null-distribution for an arbitrary predefined test statistic $T(\y,\yhat,\c)$ by sampling permutation based "copies" of $\c$,

@@

where, $Q(.|y)$ denotes the conditional distribution of $\c$ given $\y=y$ and $j=1,\dots, m$ indexes the "copy" of $\c$ so that $\c^{(j)} = (c_1^{(j)}, \dots, c_n^{(j)})$ is a permutation of the original vector $\c = (c_1, \dots, c_n)$.

This mechanism creates copies $\c^{(1)}, \dots ,\c^{(m)}$ so that under the null hypothesis ($\yhat \independent \c | \y$), the triples $$(\y,\yhat,\c), (\y, \yhat, \c^{(1)}),\dots, (\y, \yhat, \c^{(m)})$$ are all identically distributed and exchangeable, and so are the 
$$T(\y,\yhat,\c), T(\y, \yhat, \c^{(1)}),\dots,T(\y, \yhat, \c^{(m)})$$
test statistics, as well.

Let us denote the permutation-based copies of

$$\c^{(j)} = \c_{\pi^{(j)}}, \quad \pi \in S_n$$

where $S_n$ denote the set of all permutations on the indices $\{1,\dots,n\}$ and $\c_\pi = (c_{\pi_1}, \dots, c_{\pi_n})$ the vector $\c$ with its elements reordered according to the permutation .

In the partial confoudner test, the $\c_{\pi^{(j)}}$ copies are drawn so that:

@@

While the required permutations could be theoretically sampled with a simple Metropolis-Hastings algorithm that draws uniformly from $S_n$ at random, this way the acceptance ratio would be extremely low, even for moderate $n$ (except there is very low dependence of $\c$ on $\y$), resulting in slow mixing times. The partial confounder test can be, however, efficiently implemented with the parallelized pairwise Markov-Chain Monte Carlo sampler of \cite{berrett2020conditional} (Algorithm 1), that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data. In their Theorem 2, \cite{berrett2020conditional} verify that the resulting Markov Chain yields the desired stationary distribution, even if the number of steps is small.

@@

Obtaining a relatively accurate, independent estimate of $D(\cdot|\y)$ (of any shape) for CPT inference is important. Berrett and colleagues recommend to use a large independent sample to obtain the log-likelihood matrix of the conditional distribution $D(\cdot|Z)$, which might be often viable in semi-supervised learning settings where unlabelled data $(X, Z)$ are easier to obtain than labelled data $(X, Y, Z)$.
Since in the proposed predictive modelling, normality might be often a reasonable assumption for $\y$ and $\c$ (but not for $\yhat$), it is possible to fall back to Gaussian assumptions and thereby eliminate the need for a large, independent sample to estimate $D(\cdot|\y)$.

@@

\begin{equation}
    \label{eq:linreg}
    \c = \alpha + \beta \y + \boldsymbol{e}
\end{equation}

with the assumption that the residual error follows a Gaussian distribution. If we write $\boldsymbol{\mu} = \alpha + \beta \y$ and $\sigma^2 = E(\boldsymbol{e}^2)$ then, for the conditional Normal distribution $ (\c|\y=y_i) \sim \mathcal{N}\{\mu_i, \sigma^2\}$ and the log-likelihood, that is to be used in Eq. \ref{eq-pperm}, can be computed as follows:

@@

A straightforward choice of test statistic with the proposed Gaussian likelihood estimation is the coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification):

@@

$$T(\y, \yhat, \c^{(j)}) = R^2(\yhat, \c^{(j)})$$

@@

Using CPT to test confounder-bias in predictive modelling allows relaxing assumptions on $\yhat$ (for both the partial and the full test) but - in line with the "no free lunch" theorem, requires knowing - or putting assumptions on - the joint distribution of the other two variables ($\yhat$ and $\c$). 
\cite{berrett2020conditional} give a detailed analysis of the robustness of their CPT approach when estimating the conditional distribution with re-using the tested data and, also, against misspecifying the conditional distribution of interest.

Here I extend their results by performing simulations for estimating the power of the test in a form that is accessible for power calculations in predictive modelling (considering various weights of the confounder and the target signals in $\yhat$ and in the case of varying levels of covariance between $\y$ and $\c$).
Moreover, I perform a thorough investigation of the robustness of the test against the violation of normality (for both $(\y, \c)$ and $\yhat$).

@@

\subsubsection*{Gaussian Simulations}
The target variable $\y$ and the confounder variable $\c$ were drawn randomly from a multivariate normal distribution:

$$ \y, \c \sim \mathcal{N}(\mu_0, \Sigma_\sigma) $$

@@

Next, the simulated predicted values are constructed:

$$ \yhat = w_y \y + w_c \c + w_e \boldsymbol{e}$$

so that $w_y + w_c + w_e = 1$, where $\boldsymbol{e} \sim \mathcal{N}(0,1)$ is the noise component.

To test the implementation for categorical variables, simulated $\y$, $\yhat$ and $\c$ variables are binarized by thresholding at 0.

@@

$$ w_{c/y} = w_c / w_y $$

@@

$$ w_{c+y} = w_c + w_y $$

@@

$$\boldsymbol{x}' = sinh(\delta sinh^{-1}(\boldsymbol{x}) - \epsilon)$$

@@

In the simulation procedure for the 'partial' confound test, 100 repetitions were performed of all combination of the following parameter values: 
\begin{itemize}
    \item covariance of $\c$ and $\y$: $\sigma \in \{0, 0.2, 0.4, 0.6, 0.8\}$
    \item confound-to-target: $w_{c/y} \in \{0, 0.1, 0.3, 0.5, 1\}$
    \item signal-in-prediction: $w_{c+y} \in \{0, 0.3, 0.6, 0.9\}$
    \item sample size: $n \in \{50, 100, 500, 1000\}$
\end{itemize}

For testing 'full' confounder bias, identical parameters were used, except that instead of the 'confound-to-target' parameter $ w_{c/y}$ we used the 'target-to-confound' parameter $ w_{y/c} = w_y/ w_c $, but with the same parameter values $w_{y/c} \in \{0, 0.1, 0.3, 0.5, 1\}$.

Next to the normally distributed case, four \emph{sinh-arcsinh} transformed cases were also investigated, with $\delta \in \{1, 1.05, 1.5, 5\}$ $\epsilon \in \{1, 3, 5, 10\}$

The 'partial' and 'full' confound tests, as implemented in version 0.9.0 of the package \emph{'mlconfound'} were run with default parameters, that is 1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations and by implying categorical variables, where needed.

@@

'Partial' and 'full' confound testing was performed with default parameters (1000 permutations, 50 Markov-chain Monte Carlo steps) as implemented in version 0.9.0 of the package \emph{'mlconfound'}.

@@

Human Connectome Project contains imaging and behavioral data of approximately 1200 healthy subjects \citep{van2013wu}. Preprocessed resting state fMRI connectivity data (partial correlation matrices) \citep{glasser2013minimal} as published with the HCP1200 release (N=999 participants with functional connectivity data) were used to build models that predict individual fluid intelligence scores (IQ), measured with Penn Progressive Matrices \citep{duncan2000neural}.

@@

Features (functional connectivities across 100 group-independent component analysis based regions) were either (i) considered in their raw form or were subject to confound mitigation approaches by (ii) feature regression \citep{rao2017predictive} or (iii) COMBAT \citep{johnson2007adjusting, fortin2018harmonization}.
The feature mitigation strategies were separately applied for acquisition batch and age group as confounder variable.

The total of 5 types of features (raw, regressing out acquisition batch, regressing out age group, COMBAT with acquisition batch, COMBAT with age group) were independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the individual fluid intelligence scores with a ridge-regression \citep{hoerl1970ridge}. The $\alpha$ parameter of the ridge model was considered as a hyperparameter ($\alpha \in \{0.00001, 0.0001, 0.001, 0.01, 0.1, 1, 10, 100, 1000, 10000, 100000\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with mean squared error as optimization metric. Confound mitigation was performed inside of the outer cross-validation loop, to avoid leakage.

Analysis code is available as jupyter notebook\footnote{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_hcp.ipynb} in the github repository of the 'mlconfound' package.

@@

The proposed tests were applied to provide evidence of motion-bias in diagnostic predictive models of autism spectrum disorder (ASD), trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} involving 866 participants; ASD: 402, healthy control (HC): 464. Preprocessed regional timeseries data was obtained as shared with the paper \citep{dadi2019benchmarking} which was based on preprocessed image data provided by the Preprocessed Connectome Project \citep{craddock2013neuro}.

@@

The total of five types (raw, feature regression of site, feature regression of motion, COMBAT with site, COMBAT with motion) of features were independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the diagnosis (DX: ASD vs. HC) with a L2-regularized logistic regression (as e.g. in \citep{dadi2019benchmarking}). The $C$ parameter of the model was considered as a hyperparameter ($C \in \{0.1, 1, 10\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with area under the receiver operator curve (AUC under ROC) as optimization metric. Confound mitigation was performed inside of the outer cross-validation loop, to avoid leakage.

Confounder testing was performed on the class probabilities (instead of the predicted labels).

@@

The conditional permutation based 'partial' and 'full' confound tests for predictive modelling have been implemented in the python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}.

@@

Simulations based on normally distributed data justified that the proposed confounder tests - in line with theory - provide a valid type I error control (For detailed results, see first row and first column on Figure \ref{fig:sim-normal} for the 'partial' test on numerical data and Supplementary Figures \ref{fig:sim-ccc-full}, \ref{fig:sim-ccb-partial}, \ref{fig:sim-ccb-full}, \ref{fig:sim-bbc-partial}, \ref{fig:sim-bbc-full}, \ref{fig:sim-bbb-partial} and \ref{fig:sim-bbc-full} for the partial test categorical data and the full tests on both numeric and categorical data). The higher 95\% confidence limit for the type I error rates for one simulation case (with 100 repetitions) is 0.11. If the test is valid, five simulation instances are expected to pass this limit across 200 H0 simulations (all null hypotheses for all simulations depicted on the same figure). The number of simulation cases with type I error rate greater than 0.11 was always smaller then this expected value.
The results suggest, moreover, that the proposed tests are somewhat overly conservative if the model has a high predictive performance (type I error rates closer to 0 than expected).

Simulations with non-normal data showed that the proposed methods are remarkably robust to the violation of normality assumptions for the target variable $\y$ and the confounder $\c$ and, as expected - completely non-parametric regarding the distribution of the predictions $\yhat$). The six simulation cases of normality violation, shown on Figure \ref{fig:sim-non-normal} involved various \emph{sinh-arcsinh} distributions, form totally normal (Fisher's kurtosis: 0, skewness) to severely skewed and heavy tailed (kurtosis: 42, skewness: 6).
The results show that the violation of normality - as expected - increases the type I error rate, but the increase - depending on the application - might not be severely problematic up to a kurtosis and skewness values as high as 5 and 2 respectively. Furthermore, normality violation becomes less problematic in case of high predictive performance (high association between the predicted and observed values).

@@

When trained on the raw (unadjusted) connectivity features, in all four cases, the confounders were statistically significantly associated to both the target variable and the model predictions (with weak-to-moderate effect sizes) and the proposed partial confounder test provided significant evidence against the null-hypothesis of no confounder bias (i.e. conditional independence between the confounder and the predictions, given the target).
